### 2017-03-30 paper/abstract.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\begin{abstract}
Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes. In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive. In this paper we will use this concept to express parallelism.
\\\\
First, we give an introduction to some of the possible ways to add parallelism to Haskell programs. Then, we give the basic definition of Arrows, which is followed up by the introduction of some utility functions used in this paper. Next, we introduce the \code{ArrowParallel} typeclass together with backends for it written with the parallel Haskells introduced earlier, finishing up with the benefits of this new way of writing parallel programs. After this we give the definition of several parallel skeletons. Then, we introduce some syntactic sugar to mimic the sequential arrow combinators introduced by Hughes \citHughes ~but with parallelism added. We also give benchmarks of our newly created parallel Haskell. Finally we give a short conclusion of what we managed to achieve.
\end{abstract}

>>>>

\begin{abstract}
Arrows form a general interface for computation and pose therefore as an alternative to monads for API design. In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive. We express parallelism using this concept. We define an Arrows-based interface for parallel computations and implement it using multiple parallel Haskells.

This new way of writing parallel programs has a benefit of being portable across flavours of parallel Haskells used. With our framework we define  several parallel skeletons. We also  introduce some syntactic sugar to provide parallelism-aware arrow combinators similar to the traditional ones. Benchmarks shows that our framework does not induce too much overhead performance-wise.
\end{abstract}

### 2017-03-30 paper/arrows.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\section{Arrows}
\begin{center}
	\includegraphics[]{images/arrow.png}
\end{center}
Arrows were introduced by Hughes \citHughes ~as a general interface for computation. An arrow \code{arr a b} can be look at as a computation that converts an input \code{a} to an output \code{b}. This is defined in the arrow typeclass:

@@

first :: arr a b -> arr (a,c) (b,c)
\end{lstlisting}
\end{minipage}
~~~~
\begin{minipage}{0.25\textwidth}
	~\\~\\~\\
	\includegraphics[scale=0.6]{images/arr}~\\~\\
	\includegraphics[scale=0.6]{images/compose}~\\~\\
	\includegraphics[scale=0.6]{images/first}~\\~\\
\end{minipage}
\end{minipage}
\lstinline{arr} is used to lift an ordinary function to an arrow type. This can be thought of as analogous to the monadic \lstinline{return}. The \lstinline{>>>} operator, in a similar way, is analogous to the monadic composition operator \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}. And lastly, the \lstinline{first} operator, which takes the input arrow from b to c and converts it into an arrow on pairs with the second argument untouched, is also needed for actual useful code as without it, we wouldn't have a way to save input across arrows.
\\\\
The most prominent instances of this interface are regular functions \lstinline{(->)},
\begin{lstlisting}[frame=htrbl]
instance Arrow (->) where
	arr f = f
	f >>> g = g . f
	first f = \(a, c) -> (f a, c) 
\end{lstlisting}
and the Kleisli type
\begin{lstlisting}[frame=htrbl]
data Kleisli m a b = Kleisli { run :: a -> m b }
\end{lstlisting}
as well:
\begin{lstlisting}[frame=htrbl]
instance Monad m => Arrow (Kleisli m) where
	arr f = Kleisli $ return . f
	f >>> g = Kleisli $ \a -> f a >>= g
	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
\end{lstlisting}
\begin{figure}
	\centering
	\begin{tabular}{cc}
		\subcaptionbox{first\label{1}}{\includegraphics[width = 1.5in]{images/first}} &
		\subcaptionbox{second\label{2}}{\includegraphics[width = 1.5in]{images/second}} \\
		\subcaptionbox{(***)\label{1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		\subcaptionbox{(\&\&\&)\label{2}}{\includegraphics[width = 1.5in]{images/andandand}}\
	\end{tabular}
	\caption{syntactic sugar}
	\label{3}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},
\begin{lstlisting}[frame=htrbl]
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
the *** combinator which combines first and second to handle two inputs in one arrow,
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{lstlisting}
and the \&\&\& combinator that constructs an arrow which outputs 2 different values like ***, but takes only one input.
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{lstlisting}
A short example given by Hughes on how to use this is \lstinline{add} over arrows:
\begin{lstlisting}[frame=htrbl]
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
\\\\
\textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used. We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}

>>>>

\subsection{Arrows}
\label{sec:arrows}
\begin{center}
	\includegraphics[]{images/arrow.png}
\end{center}
Arrows were introduced by Hughes \citHughes\ as a general interface for computation. An arrow \code{arr a b} represents  a computation that converts an input \code{a} to an output \code{b}. This is defined in the arrow typeclass:

@@

first :: arr a b -> arr (a,c) (b,c)
\end{lstlisting}
\end{minipage}
~~~~
\begin{minipage}{0.25\textwidth}
	~\\~\\~\\
	\includegraphics[scale=0.6]{images/arr}~\\~\\
	\includegraphics[scale=0.6]{images/compose}~\\~\\
	\includegraphics[scale=0.6]{images/first}~\\~\\
\end{minipage}
\end{minipage}
\lstinline{arr} is used to lift an ordinary function to an arrow type, similarly to the monadic \lstinline{return}. The \lstinline{>>>} operator is analogous to the monadic composition  \lstinline{>>=} and combines two arrows \code{arr a b} and \code{arr b c} by "wiring" the outputs of the first to the inputs to the second to get a new arrow \code{arr a c}. Lastly, the \lstinline{first} operator  takes the input arrow from \lstinline{b} to \lstinline{c} and converts it into an arrow on pairs with the second argument untouched. It allows us to to save input across arrows.
\\\\
The most prominent instances of this interface are regular functions \lstinline{(->)},
\begin{lstlisting}[frame=htrbl]
instance Arrow (->) where
	arr f = f
	f >>> g = g . f
	first f = \(a, c) -> (f a, c) 
\end{lstlisting}
and the Kleisli type
\begin{lstlisting}[frame=htrbl]
data Kleisli m a b = Kleisli { run :: a -> m b }

instance Monad m => Arrow (Kleisli m) where
	arr f = Kleisli $ return . f
	f >>> g = Kleisli $ \a -> f a >>= g
	first f = Kleisli $ \(a,c) -> f a >>= \b -> return (b,c)
\end{lstlisting}%$
\begin{figure}
	\centering
	\begin{tabular}{cc}
		\subcaptionbox{first\label{t1}}{\includegraphics[width = 1.5in]{images/first}} &
		\subcaptionbox{second\label{t2}}{\includegraphics[width = 1.5in]{images/second}} \\
		\subcaptionbox{(***)\label{c1}}{\includegraphics[width = 1.5in]{images/starstarstar}} &
		\subcaptionbox{(\&\&\&)\label{c2}}{\includegraphics[width = 1.5in]{images/andandand}}\
	\end{tabular}
	\caption{Syntactic sugar for arrows. \olcomment{FIXME: subreferences}}
	\label{3}
\end{figure}
With this typeclass in place, Hughes also defined some syntactic sugar: The mirrored version of \lstinline{first}, called \lstinline{second},
\begin{lstlisting}[frame=htrbl]
second :: Arrow arr => arr a b -> arr (c, a) (c, b)
second f = arr swap >>> first f >>> arr swap
	where swap (x, y) = (y, x)
\end{lstlisting}
the *** combinator which combines first and second to handle two inputs in one arrow,
\begin{lstlisting}[frame=htrbl]
(***) :: Arrow arr => arr a b -> arr c d -> arr (a, c) (b, d)
f *** g = first f >>> second g
\end{lstlisting}
and the \&\&\& combinator that constructs an arrow which outputs 2 different values like ***, but takes only one input.
\begin{lstlisting}[frame=htrbl]
(&&&) :: Arrow arr => arr a b -> arr a c -> a a (b, c)
f &&& g = arr (\a -> (a, a)) >>> (f *** g)
\end{lstlisting}
A short example given by Hughes on how to use this is \lstinline{add} over arrows:
\begin{lstlisting}[frame=htrbl]
add :: Arrow arr => arr a Int -> arr a Int -> arr a Int
add f g = (f &&& g) >>> arr (\(u, v) -> u + v)
\end{lstlisting}
% The benefit of using the \code{Arrow} typeclass is that any type which is shown to be an arrow can be used in conjunction with this newly created \lstinline{add} combinator. Even though this example is quite simple, the power of the arrow interface immediately is clear: If a type is an arrow, it can automatically used together with every library that works on arrows. Compared to simple monads, this enables us to write code that is more extensible, without touching the internals of the specific arrows.
% \\\\
% \textit{Note: In the definitions Hughes gave in his paper, the notation \code{a b c} for an arrow from \code{b} to \code{c} is used. We use the equivalent definition \code{arr a b} for an arrow from \code{a} to \code{b} instead, to make it easier to find the arrow type in type signatures.}
%
The more restrictive interface of arrows (a monad can be \emph{anything}, an arrow is a process of doing something, a \emph{computation}) allows for more elaborate composition and transformation combinators. One of the major problems in parallel computing is composition of parallel processes.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-03-30 paper/basicSkeletons.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\section{Basic Skeletons}

With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some basic parallel skeletons.

\subsection{parEvalNLazy}
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalNLazy}
\end{center}
\code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. \code{map (arr . (+)) [1..]} or just because we need the arrows evaluated in chunks. \code{parEvalNLazy} fixes this. It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}. These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}. The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
\begin{lstlisting}[frame=htrbl]
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
	listApp fchunks >>>
	arr concat
	where
		fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
\end{lstlisting}

\subsection{parEval2}
\begin{center}
	\includegraphics[scale=0.7]{images/parEval2}
\end{center}
We have only talked about the paralellization arrows of the same type until now. But sometimes we want to paralellize heterogenous types as well. However, we can implement such a \code{parEval2} combinator which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass. The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
\\\\
We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}. Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above. After the calculation is finished we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}. The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
\begin{lstlisting}[frame=htrbl]
parEval2 :: (ArrowChoice arr,
	ArrowParallel arr (Either a c) (Either b d) conf) =>
	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
parEval2 conf f g = 
	arr Left *** (arr Right >>> arr return) >>>
	arr (uncurry (:)) >>>
	parEvalN conf (replicate 2 (f +++ g)) >>>
	arr partitionEithers >>>
	arr head *** arr head
\end{lstlisting}

>>>>

\subsection{Extending the Interface}
\label{sec:extending-interface}
With the \code{ArrowParallel} typeclass in place and implemented, we can now implement some further basic parallel interface functions. These are algorithmic skeletons that, however, mostly serve as a foundation to further, more specific algorithmic skeletons.

\subsection{Lazy \code{parEvalN}}
\begin{center}
	\includegraphics[scale=0.7]{images/parEvalNLazy}
\end{center}
The function \code{parEvalN} is 100\% strict, which means that it fully evaluates all passed arrows. Sometimes this might not be feasible, as it will not work on infinite lists of functions like e.g. \code{map (arr . (+)) [1..]} or just because we need the arrows evaluated in chunks. \code{parEvalNLazy} fixes this. It works by first chunking the input from \code{[a]} to \code{[[a]]} with the given \code{ChunkSize} in \code{arr (chunksOf chunkSize)}. These chunks are then fed into a list \code{[arr [a] [b]]} of parallel arrows created by feeding chunks of the passed \code{ChunkSize} into the regular parEvalN by using \code{listApp}. The resulting \code{[[b]]} is lastly converted into \code{[b]} with \code{arr concat}.
\begin{lstlisting}[frame=htrbl]
parEvalNLazy :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> [arr a b] -> (arr [a] [b])
parEvalNLazy conf chunkSize fs =
	arr (chunksOf chunkSize) >>>
	listApp fchunks >>>
	arr concat
	where fchunks = map (parEvalN conf) $ chunksOf chunkSize fs
\end{lstlisting} %$ %% formatting

\subsection{Heterogenous tasks}
\begin{center}
	\includegraphics[scale=0.7]{images/parEval2}
\end{center}
We have only talked about the paralellization arrows of the same type until now. But sometimes we want to paralellize heterogenous types as well. However, we can implement such a \code{parEval2} combinator which combines two arrows \code{arr a b} and \code{arr c d} into a new parallel arrow \code{arr (a, c) (b, d)} quite easily with the help of the \code{ArrowChoice} typeclass. The idea is to use the \code{+++} combinator which combines two arrows \code{arr a b} and \code{arr c d} and transforms them into \code{arr (Either a c) (Either b d)} to get a common arrow type that we can then feed into parEvalN.
\\\\
We start by transforming the \code{(a, c)} input into a 2-element list \code{[Either a c]} by first tagging the two inputs with \code{Left} and \code{Right} and wrapping the right element in a singleton list with \code{return} so that we can combine them with \code{arr (uncurry (:))}. Next, we feed this list into a parallel arrow running on 2 instances of \code{f +++ g} as described above. After the calculation is finished we convert the resulting \code{[Either b d]} into \code{([b], [d])} with \code{arr partitionEithers}. The two lists in this tuple contain only 1 element each by construction, so we can finally just convert the tuple to \code{(b, d)} in the last step.
\begin{lstlisting}[frame=htrbl]
parEval2 :: (ArrowChoice arr,
	ArrowParallel arr (Either a c) (Either b d) conf) =>
	conf -> arr a b -> arr c d -> arr (a, c) (b, d)
parEval2 conf f g = 
	arr Left *** (arr Right >>> arr return) >>>
	arr (uncurry (:)) >>>
	parEvalN conf (replicate 2 (f +++ g)) >>>
	arr partitionEithers >>>
	arr head *** arr head
\end{lstlisting}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-03-30 paper/conclusion.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\section{Conclusion}
As we have seen in this paper, arrows are a useful tool for composing parallel programs. By basing our parallel Haskell on them, we do not have to introduce new monadic types that wrap the computation and instead can use them just like they were regular sequential pure code. Performancewise, parallel arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.

\fixme{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}

>>>>

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition combinators. To our knowledge we are the first ones to represent parallel computation with arrows.\olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work}

Arrows turn out to be a useful tool for composing in parallel programs. We do not have to introduce new monadic types that wrap the computation. Instead use arrows just like regular sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows feature an implementation of the \code{ArrowParallel} interface for Multicore Haskell, \code{Par} monad, and Eden. With our approach parallel programs can be ported across these flavours with no effort.
%
%
Performancewise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead.\olcomment{PROOFS. Many proofs in benchmarks!}

\mbcomment{ArrowApply (or equivalent) are needed because we basically want to be able to produce intermediary results, this is by definition of the parallel evaluation combinators}

\olcomment{Remove websites from citations, put them into footnotes!}

### 2017-03-30 paper/futures.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\section{Futures} \label{futures}
Consider the following arbitrary, but interesting, parallel arrow combinator:
\begin{lstlisting}[frame=htrbl]
someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
\end{lstlisting}
In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node. While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in figure \ref{fig:withoutFutures}. This can become a serious bottleneck in heavy threaded applications.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
	\caption[without Futures]{Communication between 4 threads without Futures}
	\label{fig:withoutFutures}
\end{figure}
To fix this, we have to introduce a way that enables the nodes to communicate directly with each other. Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour. But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \code{Future} typeclass:
\begin{lstlisting}[frame=htrbl]
class Future fut a | a -> fut where
    put :: (Arrow arr) => arr a (fut a)
    get :: (Arrow arr) => arr (fut a) a
\end{lstlisting}
As \code{RD} is only type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though:
\begin{lstlisting}[frame=htrbl]
data RemoteData a = RD { rd :: RD a }

@@

instance (NFData a) => Future BasicFuture a where
    put = arr (\a -> BF { val = a })
    get = arr val
\end{lstlisting}
To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
We need this implementation for our \code{RemoteData} wrapper
\begin{lstlisting}[frame=htrbl]
instance NFData (RemoteData a) where
    rnf = rnf . rd
instance Trans (RemoteData a)
\end{lstlisting}
and the following for the \code{BasicFuture} dummy type:
\begin{lstlisting}[frame=htrbl]
instance (NFData a) => NFData (BasicFuture a) where
	rnf = rnf . val
\end{lstlisting}
Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
\begin{lstlisting}[frame=htrbl]
someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 =
	parEvalN () (map (>>> put) fs1) >>>
	rightRotate >>>
	parEvalN () (map (get >>>) fs2)
\end{lstlisting}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in figure \ref{fig:withFutures}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 threads with Futures}
	\label{fig:withFutures}
\end{figure}

>>>>

\section{Futures} \label{futures}
Consider the following  parallel arrow combinator:
\begin{lstlisting}[frame=htrbl]
someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 = parEvalN () fs1 >>> rightRotate >>> parEvalN () fs2
\end{lstlisting}
In a distributed environment, the resulting arrow of this combinator first evaluates all \code{[arr a b]} in parallel, sends the results back to the master node, rotates the input once and then evaluates the \code{[arr b c]} in parallel to then gather the input once again on the master node. 
Such situations arise, \eg in scientific computations when the data distributed across the nodes needs to be transposed. A concrete example is 2D FFT computation \cite{Gorlatch,pact-fft}.

While this could be rewritten into only one \code{parEvalN} call by directly wiring the arrows properly together, this example illustrates an important problem: When using a \code{ArrowParallel} backend that resides on multiple computers, all communication between the nodes is done via the master node, as shown in the Eden trace in Figure~\ref{fig:withoutFutures}. This can become a serious bottleneck %in heavy threaded applications.
for larger amount of data and number of processes \citep[showcases][as, \eg]{Berthold2009b}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withoutFutures}
	\caption[without Futures]{Communication between 4 threads without Futures}
	\label{fig:withoutFutures}
\end{figure}
This motivates for an approach that allows the nodes to communicate directly with each other. Thankfully, Eden, the distributed parallel Haskell we have used in this paper so far, already ships with the concept of \code{RD} (remote data) that enables this behaviour \cite{AlGo03a,dieterle2010remotedata}. But as we want code written against our API to be implementation agnostic, we have to wrap this context. We do this with the \code{Future} typeclass:
\begin{lstlisting}[frame=htrbl]
class Future fut a | a -> fut where
    put :: (Arrow arr) => arr a (fut a)
    get :: (Arrow arr) => arr (fut a) a
\end{lstlisting}
As \code{RD} is only type synonym for communication type that Eden uses internally, we have to use some wrapper classes to fit that definition, though:
\begin{lstlisting}[frame=htrbl]
data RemoteData a = RD { rd :: RD a }

@@

instance (NFData a) => Future BasicFuture a where
    put = arr (\a -> BF { val = a })
    get = arr val
\end{lstlisting}
To fit the \code{ArrowParallel} instances we gave earlier, we also have to give the necessary \code{NFData} and \code{Trans} instances - the latter only being needed in Eden.
We need this implementation for our \code{RemoteData} wrapper
\begin{lstlisting}[frame=htrbl]
instance NFData (RemoteData a) where
    rnf = rnf . rd
instance Trans (RemoteData a)
\end{lstlisting}
and the following for the \code{BasicFuture} dummy type:
\begin{lstlisting}[frame=htrbl]
instance (NFData a) => NFData (BasicFuture a) where
	rnf = rnf . val
\end{lstlisting}
Going back to our communication example we can use this Future concept in order to enable direct communications between the nodes in the following way:
\begin{lstlisting}[frame=htrbl]
someCombinator :: (Arrow arr) => [arr a b] -> [arr b c] -> arr [a] [c]
someCombinator fs1 fs2 =
	parEvalN () (map (>>> put) fs1) >>>
	rightRotate >>>
	parEvalN () (map (get >>>) fs2)
\end{lstlisting}
In a distributed environment, this gives us a communication scheme with messages going through the master node only if it is needed - similar to what is shown in the trace in Fig.~\ref{fig:withFutures}.\olcomment{Fig.~3 is not really clear. Do Figs 2-3 with a lot of load?}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.9\textwidth]{images/withFutures}
	\caption[with Futures]{Communication between 4 threads with Futures}
	\label{fig:withFutures}
\end{figure}

### 2017-03-30 paper/mapSkeletons.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\section{Map-based Skeletons}
\subsection{parMap}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = parEvalN conf (repeat f)
\end{lstlisting}

\subsection{parMapStream}
\begin{center}
	\includegraphics[scale=0.7]{images/parMapStream}
\end{center}
As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}. So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
\begin{lstlisting}[frame=htrbl]
parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> arr a b -> arr [a] [b]
parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
\end{lstlisting}

\subsection{farm}
\begin{center}
	\includegraphics[scale=0.7]{images/farm}
\end{center}
\code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper). This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf,
	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	unshuffle numCores >>>
	parEvalN conf (repeat (mapArr f)) >>>
	shuffle
\end{lstlisting}
The definition of \code{unshuffle} is
\begin{lstlisting}[frame=htrbl]
unshuffle :: (Arrow arr) => Int -> arr [a] [[a]]
unshuffle n = arr (\xs -> [takeEach n (drop i xs) | i <- [0..n-1]])

takeEach :: Int -> [a] -> [a]
takeEach n [] = []
takeEach n (x:xs) = x : takeEach n (drop (n-1) xs)
\end{lstlisting}
, while \code{shuffle} is defined as:
\begin{lstlisting}[frame=htrbl]
shuffle :: (Arrow arr) => arr [[a]] [a]
shuffle = arr (concat . transpose)
\end{lstlisting}
(These were taken from Eden's source code \cite{eden_skel_shuffle} and translated into Arrows.)

@@

\subsection{parMapReduce}
-- this does not completely adhere to Google's definition of Map Reduce as it
-- the mapping function does not allow for "reordering" of the output
-- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce

@@

>>>>

\section{Map-based Skeletons}
\label{sec:map-skeletons}
Now we have developed Parallel Arrows further enough to define some algorithmic skeletons useful to an application programmer.
\subsection{Parallel map}
\begin{center}
	\includegraphics[scale=0.7]{images/parMap}
\end{center}
\code{parMap} is probably the most common skeleton for parallel programs. We can implement it with \code{ArrowParallel} by repeating an arrow \code{arr a b} and then passing it into \code{parEvalN} to get an arrow \code{arr [a] [b]}. 
Just like \code{parEvalN}, \code{parMap} is 100 \% strict.
\begin{lstlisting}[frame=htrbl]
parMap :: (ArrowParallel arr a b conf) =>
	conf -> (arr a b) -> (arr [a] [b])
parMap conf f = parEvalN conf (repeat f)
\end{lstlisting}

\subsection{Lazy parallel map}
\begin{center}
	\includegraphics[scale=0.7]{images/parMapStream}
\end{center}
As \code{parMap} is 100\% strict it has the same restrictions as \code{parEvalN} compared to \code{parEvalNLazy}. So it makes sense to also have a \code{parMapStream} which behaves like \code{parMap}, but uses \code{parEvalNLazy} instead of \code{parEvalN}.
\begin{lstlisting}[frame=htrbl]
parMapStream :: (ArrowParallel arr a b conf, ArrowChoice arr, ArrowApply arr) =>
	conf -> ChunkSize -> arr a b -> arr [a] [b]
parMapStream conf chunkSize f = parEvalNLazy conf chunkSize (repeat f)
\end{lstlisting}

\subsection{Statically load-balancing parallel map}
\begin{center}
	\includegraphics[scale=0.7]{images/farm}
\end{center}
\code{parMap} spawns every single computation in a new thread (at least for the instances of \code{ArrowParallel} we gave in this paper). This can be quite wasteful and a \code{farm} that equally distributes the workload over \code{numCores} workers (if numCores is greater than actualProcessorCount, the fastest processor(s) to finish will get more tasks) seems useful.
\begin{lstlisting}[frame=htrbl]
farm :: (ArrowParallel arr a b conf,
	ArrowParallel arr [a] [b] conf, ArrowChoice arr) =>
	conf -> NumCores -> arr a b -> arr [a] [b]
farm conf numCores f =
	unshuffle numCores >>>
	parEvalN conf (repeat (mapArr f)) >>>
	shuffle
\end{lstlisting}

@@

\subsection{parMapReduce}
\olcomment{it appears STRANGE. are the data really left alone and noded after map and taken from there by reduce? It makes sense only is no communication through master takes place. ELSE: CUT!}

-- this does not completely adhere to Google's definition of Map Reduce as it
-- the mapping function does not allow for "reordering" of the output
-- The original Google version can be found at https://de.wikipedia.org/wiki/MapReduce

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

### 2017-03-30 paper/motivation.tex
merged with Oleg's first changes. only topologySkeletons.tex has to be merged

\section{Motivation}
Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes. In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive. In this paper we will use this concept to express parallelism.

>>>>

%\section{Motivation}
%Arrows were introduced in John Hughes paper as a general interface for computation and therefore as an alternative to monads for API design \citHughes. In the paper Hughes describes how arrows are a generalization of monads and how they are not as restrictive. In this paper we will use this concept to express parallelism.

\section{Introduction}
\label{sec:introduction}
\olcomment{todo, reuse 5.5, "Impact" at the end and more}

blablabla arrows, parallel, haskell.

\paragraph{Contribution}

HIT HERE REALLY STRONG

\paragraph{Structure}
The remaining text is structures as follows. Section~\ref{sec:background} briefly introduces known parallel Haskell flavours and gives an overview of Arrows to the reader (Sec.~\ref{sec:arrows}). Section~\ref{sec:related-work} discusses related work. Section~\ref{sec:parallel-arrows} defines Parallel Arrows and presents a basic interface. Section~\ref{futures} defines futures for Parallel Arrows. Futures are a further concept that helps to define more advanced interfaces. Section~\ref{sec:map-skeletons} presents some basic algorithmic skeletons in our newly defined dialect. More advanced ones are showcased in Section~\ref{sec:topology-skeletons}. Section~\ref{sec:benchmarks} shows the benchmark results. Section~\ref{sec:conclusion} concludes.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
